A C-to-RTL flow as an energy efficient alternative to embedded processors in digital systems

### 2010-05-31 Documents/Publications/dsd2010/paper.tex
final version of tex file.

The design of complex digital systems is expensive due to two reasons:
the need for trained manpower, and the difficulty of verification at
every step of the design process. Thus, it is often the case that
low-power embedded microprocessors are used to implement complex
algorithms in digital systems, so that the design and verification
problems are moved to the software domain. A high-level synthesis flow
which maps an algorithm directly to a hardware description can be a
potential alternative if it provides a verifiable and optimisable path
from executable specifications\cite{gajski-executable-specification}
to hardware.

@@

Our high-level synthesis flow is based on a factored intermediate
representation which guarantees the correctness of the implementation,
and supports optimisations that can scale with the size of the
program. The synthesis flow introduces an intermediate step in the
form of a representation called AHIR (A Hardware Intermediate
Representation). The representation is independent of the programming
language used, and can be routinely translated to a hardware
implementation, while also supporting scalable
optimisations\cite{AHIR_sameerds_hakim_2007}\cite{ahir_thesis}.

@@

There has been a great deal of activity in the field of 
high-level synthesis in the context of converting a high
level program to RTL.   Attempts at creating a path from 
high-level programs to hardware descriptions can be loosely 
categorised as follows:

@@

Some efforts simply interpret a program as a behavioral
specification, which is mapped to a hardware implementation using an
intermediate representation.

@@

Our work is similar, since the goal is to transparently compile
programs into hardware. AHIR differs from both Pegasus and SPARK since
it factorises the system into three separate components: control flow,
data flow and memory.  The components can be optimised and
implemented separately as long as specified constraints are satisfied.

Several commercial tools which translate a C program to hardware are
also available. For example,
Catapult-C\footnote{\url{http://www.mentor.com/products/esl/high_level_synthesis/}}
from Mentor Graphics,
C-to-Silicon\footnote{\url{http://www.cadence.com/products/sd/silicon_compiler/}}
from Cadence, and Synopsys
HLS\footnote{\url{http://www.synopsys.com/Tools/SLD/HLS/}} from
Synopsys. Currently, we have no access to the internal details of
these proprietary tools.

@@

The control-path has a single marked place in the initial marking.
This enables a single transition called \sym{init}, which responds to
a request symbol in alphabet $\Omega$ and begins execution of the
module. The end of execution is represented by the \sym{fin}
transition that emits an acknowledgment symbol in $\Omega$ and marks
the initially marked place.

@@

Control paths in AHIR are instances of Type-2 petri-nets (introduced
in \cite{ahir_thesis}), which
are defined using a set of standard construction rules. 
The structure of a Type-2 petri-net is
designed to enable analyses and transformations that are scalable with
the size of the petri-net.

@@

We describe an optimization that reuses a data-path operator for
multiple operations that are never active at the same time. This
avoids arbitration overheads since there is no contention. The
optimisation uses an almost linear static analysis of the control-path
to identify sharing opportunities.

@@

\begin{definition} Two elements $e_1$ and $e_2$ in a Type-2 petri-net
are compatible if and only if one of the following is true:
\begin{enumerate}
\item their NCA is {\em not} a fork region
\item there is a path within the NCA fork region, joining the two elements
\end{enumerate}
\end{definition}

@@

The compatibility label is a record of every path reaching that
element from the $init$ transition, which results in an exponential
size. Comparing two labels for compatibility also has exponential
complexity, since every sequence in one label has to be compared with
every sequence in the other label.

We eliminate the complexity by using a {\em label representation
  graph} (LRG) as shown in Fig.~\ref{figure:LRG}. The LRG represents
labels as nodes, where edges represent the manner in which a label is
constructed from other labels. The LRG is a directed acyclic graph
with a single root node that represents the empty label.

@@

We have used the LRG to implement arbiterless sharing in the AHIR
synthesis flow. We use a simple greedy algorithm to generate cliques
of pair-wise compatible operations that are mapped to a single
operator in the data-path. The incoming data-edges are routed through
multiplexers; the registers for the outgoing data-edges are not
shared.

This scheme for arbiterless sharing is quite effective in reducing
hardware costs, demonstrating support for scalable optimisations in
AHIR. Synthesis results show improvements in the throughput-area ratio
(measured in $\operatorname{Hz}/\operatorname{slice}$) in the range of
15--190\% depending on the application\cite{ahir_thesis}.

@@

The synthesis flow uses the LLVM Compiler
Infrastructure\footnote{\url{http://llvm.org/}} to parse and optimise
the input C program. This is translated to an AHIR specification using
a CDFG as an intermediate step.

@@

The LLVM bytecode is then translated to a control data flow graph
(CDFG)\cite{conditional_branches_rim_jain_1992}-%
\cite{functional_synthesis_TASS_amellal_kaminska_1994}, as shown in
Fig.~\ref{fig:snippet-cdfg}. The CDFG represents instructions as
nodes connected by control and data flow edges. The edges in the CDFG
arise from three kinds of dependence in the original program:

\begin{itemize}
\item Data dependence that creates control and data flow.
\item Control structures that create control flow.
\item External dependence (in storage) that creates control flow.
\end{itemize}

@@

The conversion of an AHIR network to VHDL is straightforward.
Currently, we produce a system which is synchronous and uses
a single clock (positive edges only).  Each symbol that is
exchanged between the control path and the data path is coded 
by a pulse which is sampled high by exactly one clock
edge.  Symbols are associated with transitions as indicated
earlier.

Places in the AHIR control path are handled as follows:
\begin{enumerate}
\item If a place has a single predecessor transition which is
an output transition and a single successor which is an input
transition, then it is optimized away.
\item If a place has a single successor transition which is
an output transition, and if this successor transition is
not a join, then the place is modeled by an OR gate whose inputs
are its predecessor transitions.
\item In the rest of the cases, the place is modeled as 
a finite state machine with two states, with the state being
set by any of its predecessors, and reset by any of its successors.
\end{enumerate}
The number of places that need to be modeled by state machines is
usually a small fraction of the total number of places.
Output transitions are modeled by AND gates, with each predecessor place
to the transition providing an input to the AND gate.  The cost
of implementing the control path is thus minimal in relation to
the implementation of the data path and memory subsystem.

Operators in the data path compute their results in a single
clock cycle, and the outputs are registered.  Shared operators
are constructed using multiplexers in a standard manner.
The instantiation of operators in the data path, 
and the interconnections within
the data path and between the data path and the control path
mirror those in the AHIR representation, except for the fact
that several operators in the AHIR data path can be mapped
to the same physical operator (as identified by arbiterless
operating sharing).
The inter module link layer is completed using the call graph
corresponding to the source code.

@@

As a reference to judge the energy efficiency of the AHIR generated
circuits, we consider a popular low power processor (henceforth
denoted by $P$), namely the Intel Atom N270
processor\footnote{\url{http://www.intel.com/products/processor/atom/index.htm}},
which is built in Intel's $45nm$ technology. This processor is readily
available, combines high performance with low power dissipation, has
full floating point support, and is a part of many low power computing
devices. The sample set of C programs was run on the processor $P$
(exactly the same program used in the hardware generation was run on
the processor). In each case, the number of cycles needed to finish
one task (as described above) specific to that program was measured.
The data sets for the programs were small enough to fit into the
processor cache, and the cycle count was averaged across multiple
runs. These cycle counts were then used to find the delay and energy
consumption for the processor. All values for area, frequency and
power dissipation for the processor were obtained from the
corresponding data-sheet.

@@

If we compare the results for the $180nm$ AHIR circuits with
the $45nm$ processor $P$, we find that except for the AES and
RBT cases,
the energy consumed per task by the AHIR circuit is appreciably
lower than that for the processor $P$.  This is significant
because the processor $P$ is implemented in a technology which
is four generations ahead of the AHIR circuits.

>>>>

The design of complex digital systems is expensive due to two reasons:
the need for trained manpower, and the difficulty of verification at
every step of the design process. Thus, it is often the case that
low-power embedded microprocessors are used to implement complex
algorithms in digital systems, so that the design and verification
problems are moved to the software domain. A high-level synthesis flow
which maps an algorithm directly to a hardware description can be a
potential alternative if it provides a verifiable and optimizable path
from executable specifications\cite{gajski-executable-specification}
to hardware.

@@

Our high-level synthesis flow is based on a factored intermediate
representation which guarantees the correctness of the implementation,
and supports optimizations that can scale with the size of the
program. The synthesis flow introduces an intermediate step in the
form of a representation called AHIR (A Hardware Intermediate
Representation). The representation is independent of the programming
language used, and can be routinely translated to a hardware
implementation, while also supporting scalable
optimizations\cite{AHIR_sameerds_hakim_2007}\cite{ahir_thesis}.

@@

There have been several approaches to the creation of a path from 
high-level programs to hardware descriptions, which
can be loosely categorized as follows:

@@

Some efforts interpret a program as a behavioral
specification, which is mapped to a hardware implementation using an
intermediate representation.

@@

Our work is similar, since the goal is to transparently compile
programs into hardware. AHIR differs from both Pegasus and SPARK since
it factorizes the system into three separate components: control flow,
data flow and memory.  The components can be optimized and
implemented separately as long as specified constraints are satisfied.

Several commercial tools which translate a C program to hardware are
also available, such as 
Catapult-C\footnote{\url{http://www.mentor.com/products/esl/high_level_synthesis/}}
from Mentor Graphics,
C-to-Silicon\footnote{\url{http://www.cadence.com/products/sd/silicon_compiler/}}
from Cadence, and Synopsys
HLS\footnote{\url{http://www.synopsys.com/Tools/SLD/HLS/}} from
Synopsys. Currently, we have no access to the internal details of
these proprietary tools.

@@

The control-path has a single marked place in the initial marking.
This enables a single transition called {\em init}, which responds to
a request symbol in alphabet $\Omega$ and begins execution of the
module. The end of execution is represented by the {\em fin}
transition that emits an acknowledgment symbol in $\Omega$ and marks
the initially marked place.

@@

Control paths in AHIR are instances of Type-2 petri-nets (introduced
in \cite{ahir_thesis}), which
are defined using a set of standard construction rules, which
ensure that the resulting petri-net is live and safe. 
Further, the structure of a Type-2 petri-net  
enables analyses and transformations that are scalable with
the size of the petri-net.

@@

We describe an optimization that reuses a data-path operator for
multiple operations that are never active at the same time. This
avoids arbitration overheads since there is no contention. The
optimization uses an almost linear static analysis of the control-path
to identify sharing opportunities.

@@

\begin{definition} Two elements $e_1$ and $e_2$ in a Type-2 petri-net
are compatible if their NCA is {\em not} a fork region or if
there is a directed path within the NCA region which joins the two elements.
\end{definition}

@@

The compatibility label is a record of every path reaching that
element from the $init$ transition, and can thus have an exponential
size (the number of elements in the label set can be
exponential in the size of the petri-net). 
Comparing two labels for compatibility can also have exponential
complexity, since every sequence in one label has to be compared with
every sequence in the other label.

We eliminate the exponential complexity by using a {\em label representation
  graph} (LRG) as shown in Fig.~\ref{figure:LRG}. The LRG represents
labels as nodes, where edges represent the manner in which a label is
constructed from other labels. The LRG is a directed acyclic graph
with a single root node that represents the empty label.

@@

We have used the LRG to implement arbiter-less sharing in the AHIR
synthesis flow. We use a simple greedy algorithm to generate cliques
of pair-wise compatible operations that are mapped to a single
operator in the data-path. The incoming data-edges are routed through
multiplexers; the registers for the outgoing data-edges are not
shared.

This scheme for arbiter-less sharing is quite effective in reducing
hardware costs, demonstrating support for scalable optimizations in
AHIR. Synthesis results (targeted at field programmable gate arrays)
show improvements in the throughput-area ratio
(measured in $\operatorname{Hz}/\operatorname{slice}$) in the range of
15--190\% depending on the application\cite{ahir_thesis}.

@@

The synthesis flow uses the LLVM Compiler
Framework\footnote{\url{http://llvm.org/}} to parse and optimize the
input C program. The resulting optimized LLVM bytecode is then
translated to an AHIR specification using a CDFG as an intermediate
step.

@@

The LLVM bytecode is then translated to a control data flow graph
(CDFG)\cite{conditional_branches_rim_jain_1992}-%
\cite{functional_synthesis_TASS_amellal_kaminska_1994}, as shown in
Fig.~\ref{fig:snippet-cdfg}. The CDFG is a hypergraph where the nodes
represent instructions and edges represent control and data flow. A
data edge is a hyperedge with a single tail and one or more heads,
that represents a value in the program, defined by the tail and used
by the heads. The control edge has a single tail and a single head,
and represents the passing of control from the tail to the head.

Each value in the program that is defined by some instruction $u$ and
used by a set of instructions $s_v$ is mapped a data edge in the CDFG
where the tail $n_u$ corresponds to the instruction $u$ and the set of
heads has a one to one correspondence with the set $s_v$. Control
edges are created in a manner that exposes instruction-level
parallelism in the program. Sequence is enforced only where the
execution of a node may affect execution of some other node. For
example, when a node $v$ uses the output of a node $u$, a control edge
is introduced to ensure that $v$ is executed only when $u$ is
finished. Similarly, control edges are used to eliminate hazards
between pairs of load instructions, store instructions and function
calls that may access the same memory location, as determined by a
memory reference analysis.

@@

The conversion of an AHIR circuit to VHDL is straightforward.
Currently, we produce a system which is synchronous and uses a single
clock (positive edges only). Each symbol exchanged between the control
path and the data path is coded by a pulse which is sampled high by
exactly one clock edge. Symbols are associated with transitions as
indicated earlier.

Places in the AHIR control path are handled as follows:
\begin{enumerate}
\item If a place has a single predecessor transition which is
an output transition and a single successor which is an input
transition, then it is optimized away.
\item If a place has a single successor transition which is
an output transition, and if this successor transition is
not a join, then the place is modeled by an OR gate whose inputs
are its predecessor transitions.
\item In all other cases, the place is modeled as a finite state
  machine with two states, with the state being set by any of its
  predecessors, and reset by any of its successors.
\end{enumerate}
The number of places that need to be modeled by state machines is
usually a small fraction of the total number of places.
Output transitions are modeled by AND gates, with each predecessor place
to the transition providing an input to the AND gate.  The cost
of implementing the control path is thus minimal in relation to
the implementation of the data path and memory subsystem.

Operators in the data path compute their results in a single clock
cycle, and the outputs are registered. Shared operators are
constructed using multiplexers in a standard manner. The instantiation
of operators in the data path, the interconnections within the data
path and those between the data path and the control path mirror those
in the AHIR representation, except for the fact that several operators
in the AHIR data path can be mapped to the same physical operator
(through arbiterless sharing). The inter module link layer is
completed using the call graph corresponding to the source code.

@@

As a reference to judge the energy efficiency of the AHIR generated
circuits, we consider a popular low power processor (henceforth
denoted by $P$), namely the Intel Atom N270
processor\footnote{\url{http://www.intel.com/products/processor/atom/index.htm}},
which is built in Intel's $45nm$ technology. This processor is readily
available, combines high performance with low power dissipation, has
full floating point support, and is a part of many low power computing
devices. The sample set of C programs was run on the processor $P$
(exactly the same program used in the hardware generation was run on
the processor). In each case, the number of cycles needed to finish
one task (as described above) was measured. The data sets for the
programs were small enough to fit into the processor cache, and the
cycle count was averaged across multiple runs. These cycle counts were
then used to find the delay and energy consumption for the processor.
All values for area, frequency and power dissipation for the processor
were obtained from the corresponding data-sheet.

@@

We find that the energy consumed per task by the $180nm$ AHIR circuits
is appreciably lower than that for the $45nm$ processor $P$ (except
for AES and RBT). This is significant because the processor $P$ is
implemented in a technology which is four generations ahead of the
AHIR circuits.
